Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Section 5.2

\subsection{A conductive semantics}

@@

An aspect of the approach of~\cite{KeilThiemannUnionIntersection} is to rewrite
these nested contracts using the De Morgan law of
Figure~\ref{fig:de-morgan-law}, eagerly opening up unions while postponing
intersection as much as possible.
This transformation flattens arbitrarily nested unions and
intersections to a union of intersections, making blame resolution more
manageable.\unsure{(Yann) find better (more precise) wording}

\begin{figure}[h]
\begin{lstlisting}[frame=none, numbers=none, language=nickel]
A @& (B @| C) (*@ $\simeq$ @*) (A @& B) @| (A @& C)
\end{lstlisting}
\caption{De Morgan law for contracts}
\label{fig:de-morgan-law}
\end{figure}

To implement this solution, \cite{KeilThiemannUnionIntersection} resorts to
specific reduction rules for unions and intersections which perform this
rewriting on the fly. The result is that the reduction rules sometimes feel
ad-hoc, and make us question the composability and extensibility of the system.
In the words of Williams, Morris, and Wadler (see
Section~\ref{sec:will-morr-wadl}): ``(\ldots) the monitoring semantics for
contracts of intersection and union types given by Keil and Thiemann are not
uniform. (\ldots) If uniformity helps composition, then special cases can hinder
composition.'' \cite{RootCauseOfBlame}. Another less important consequence is
that in Figure~\ref{fig:de-morgan-law}, the contract \nickel{A} is duplicated,
potentially causing additional work. If \nickel{B} fails, \nickel{A} will be
re-evaluated in each branch of the right hand side of
Figure~\ref{fig:de-morgan-law} in the absence of an additional sharing
mechanism.
\resolved{(Yann) Does
it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
\resolved{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

One important advantage of~\cite{KeilThiemannUnionIntersection} is to handle
user-defined contracts. We explained in Section~\ref{sec:issues-sem} why a
direct approach for higher-order intersection contracts (as done
in~\cite{RootCauseOfBlame}) is not trivial to adapt for user-defined contracts.
\cite{KeilThiemannUnionIntersection} provides two operational semantics of the
contract system: a non-deterministic one, more intuitive and proof-friendly, and
a deterministic one for actual implementation. The non-deterministic semantics
cleanly handle intersections with user-defined contracts: as each component is
checked in a separate thread, there is no direct interaction.

Let us recast the problematic example of Figure~\ref{fig:inter-flat-contracts}
in Figure~\ref{fig:inter-flat-contracts-2}). The execution now splits the
intersection into two sub-contracts that are checked separately. If one
component fails, a corresponding flag is set in a global state. In particular,
the two threads don't interact directly, and don't raise blame by themselves,
leading this example to run as expected.

\begin{figure}[h]
\begin{lstlisting}[language=nickel,title={Original program}]
let C = contracts.fromPred (fun f =>
  f 0 == 0) in
let g | (Str -> Str) @& C
      = fun x => x
in g 0
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={First component}]
(g | Str -> Str) 0
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={Second component}]
(g | C) 0
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts-2}
\end{figure}

In return, the complexity is offloaded to a separate procedure that must decide
when to raise blame. This complexity is salient in the deterministic, sequential
version~\cite[Section 5.1]{KeilThiemannUnionIntersection}. There,
the evaluation procedure has to deal again with the problem of eliding contracts
that are not \emph{compatible}~\cite[Section 5.1]{KeilThiemannUnionIntersection} (contracts are incompatible
if they are coming from a different branch and should be ignored). In practice,
the evaluation procedure maintains a global environment of contracts that is
extended each time a non-trivial contract (union, intersection or higher-order)
is opened. Then, any contract application must check the presence of the
corresponding label in the environment. What's more, each time a function with a contract attached is applied,
the whole \emph{context} (that is, the continuation of the application)
must be checked for compatibility with respect to each label of the contract
environment. Checking for compatibility involves searching the context for
contract applications.

Keil and Thiemann realize the tour de force of combining union and intersection
contracts with user-defined contracts. The price to pay, though, is that
implementing their solution in a reasonable way is far from obvious. Not only
when starting from the more abstract, non-deterministic formulation, but even
when starting from the algorithmic version of their system.
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}

>>>>

\subsection{A coinductive semantics}
\label{sec:coinductive-sem}

@@

All of this still holds true of the coinductive semantics described in
Section~\ref{sec:coinductive-sem}. Nevertheless,
\cite{KeilThiemannUnionIntersection} gives an algorithmic system which
is complete for their coinductive semantics. In a remarkable technical
tour de force, their algorithmic system is compatible with
user-defined contract (see Section~\ref{sec:flat-and-inter}).

A key aspect of the approach of~\cite{KeilThiemannUnionIntersection}
is to rewrite nested union and intersection contracts into a
disjunctive normal form using the De Morgan's law \nickel{A @& (B @|
  C)}$\simeq$\nickel{(A @& B) @| (A @& C)}. The goal is to be able to
delay the choice of branch in intersections as much as possible.

To implement contract verification, \cite{KeilThiemannUnionIntersection} resorts to
specific reduction rules for unions and intersections which perform this
rewriting on the fly. This aspect is critiqued
in~\cite{RootCauseOfBlame}:
``the monitoring semantics for
contracts of intersection and union types given by Keil and Thiemann are not
uniform. (\ldots) If uniformity helps composition, then special cases can hinder
composition.''

A cost of this approach is that the De Morgan's law
\nickel{A @& (B @| C)}$\simeq$\nickel{(A @& B) @| (A @& C)} duplicates
contract \nickel{A}, which will cause some contracts to be checked
several times. This can be an issue with user-defined contracts which
may include costly tests.

Efficiency is also affected another way: each time a function with a
contract attached is applied, the whole \emph{context} must be
traverse to check for a compatibility property.

The algorithmic system of~\cite{KeilThiemannUnionIntersection} is, on
balance, a technically impressive realization of the coinductive
semantics, though it is fairly complex and probably difficult to
implement efficiently.
